Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-07 paper/paper.tex
Added same explanation for laziness and used unicode to make code fit.

\begin{figure}
\begin{alltt}
  ;; State \(\rightarrow\) Setof State
  (define (step \(\mstate\))
    (match \(\mstate\)
      [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
       (match \(\mexp\)
         [(var\(\superscript\mlab\) \(\mvar\))
          (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
         [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
         [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
         [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
      [(co \(\msto\) \(\mkont\) \(\mval\))
       (match \(\mkont\)
         ['mt (set (ans \(\msto\) \(\mval\)))]
         [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
         [(fn\(\superscript\mlab\) \(\mval'\))
          (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
            (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
         [(fi\(\superscript\mlab\) c a \(\rho\))
          (for/set ((k (get-cont \(\msto\) l)))
            (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
      [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
       (match fun
         [(clos l \(\mvar\) \(\mexp\) \(\rho\))
          (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
          (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
         [(op \(\mop\))
          (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
                     (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [_ (set)]))]))
\end{alltt}
\caption{Implementation of machine transition relation.}
\end{figure}

@@

\begin{alltt}
;; Expr \(\rightarrow\) Setof State
(define (eval e)
  (fix step (set (inj e))))
\end{alltt}

@@

\begin{verbatim}
;; appl : (∀ (X) ((X -> (Setof X)) -> ((Setof X) -> (Setof X))))
(define ((appl f) s)
  (for/fold ([i (set)])
    ([x (in-set s)])
    (set-union i (f x))))

;; Calculate fixpoint of (appl f).
;; fix : (∀ (X) ((X -> (Setof X)) (Setof X) -> (Setof X)))
(define (fix f s)
  (let loop ((accum (set)) (front s))
    (if (set-empty? front)
        accum
        (let ((new-front ((appl f) front)))
          (loop (set-union accum front)
                (set-subtract new-front accum))))))
\end{verbatim}

@@

\begin{verbatim}
;; State^ = (cons (Set Conf) Store)

;; (State -> Setof State) -> State^ -> { State^ }
(define ((wide-step step) state)
  (match state
    [(cons cs \(\msto\))
     (define ss ((appl step)
                 (for/set ([c cs]) (c->s c \(\msto\)))))
     (set (cons (for/set ([s ss]) (s->c s))
                (join-stores ss)))]))
\end{verbatim}

@@

\begin{align*}
force(\clos{x,e,\menv},\msto) &= \{\clos{x,e,\menv}\}\\
force(\saddr\maddralt,\msto) &= \msto(\maddralt)\\
\ev{\svar{x},\menv,\mkont,\msto} &\machstep\; 
\co{\mkont,\saddr{\menv(x)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},v,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{v,\maddr}} \\
\mbox{where } &v' \in force(v,\msto)
bind^\mcntr_\mlab(\msto,x,\menv,v) =
\end{align*}
Lazy:
   cpu time: 32481 real time: 32881 gc time: 547

@@

\begin{figure}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \compile{\mexp}(\epsilon,\varnothing,\varnothing,\kmt) \multimachstep \mstate \} \text{ where }
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr},\mval,\msto} & \machstep
\mcomp^\mcntr(\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr})
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue,\msto} & \machstep
\mcompi0^\mcntr(\menv,\msto,\mkont)
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse,\msto} & \machstep
\mcompi1^\mcntr(\menv,\msto,\mkont)
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\msto,\mkont} & \machstep
\mcomp^{\mcntr'}(\menv',\msto',\mkont)
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
\\
\ap{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where }\mkont\in\msto(\maddr)
\text{ and } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for compiled ISWIM}
\label{fig:caam}
\end{figure}

@@

\begin{alltt}
;; State^ -> State^
;; Specialized from wide-step : State^ -> { State^ } ≈ State^ -> State^
(define (wide-step-specialized state)
  (match state
    [(cons \(\msto\) cs)
     (define-values (cs* \(\msto\)*)
       (for/fold ([cs* (set)] [\(\msto\)* \(\msto\)])
         ([c cs])
         (match (step-compiled^ (cons \(\msto\) c))
           [(cons \(\msto\)** cs**)
            (values (set-union cs* cs**) (join-store \(\msto\)* \(\msto\)**))])))
     (cons \(\msto\)* (set-union cs cs*))]))
\end{alltt}
%% Should be commutted to beginning

@@

\begin{alltt}
;; State^ -> State^
;; Specialized from wide-step : State^ -> { State^ } ≈ State^ -> State^
(define (wide-step-specialized state)
  (match state
    [(cons \(\msto\) cs)
     (define-values (cs* ∆)
       (for/fold ([cs* (set)] [∆* '()])
         ([c cs])
         (match (step-compiled^ (cons \(\msto\) c))
           [(cons \(\msdiff\)** cs**)
            (values (set-union cs* cs**) (append \(\msdiff\)** \(\msdiff\)*))])))
     (cons (update \(\msdiff\) \(\msto\)) (set-union cs cs*))]))
\end{alltt}

@@

\begin{verbatim}
 (let ((id (\(\lambda\) (x) x)))
   (begin (id 1) (id 2)))
\end{verbatim}

>>>>

\begin{figure}
\begin{alltt}
  ;; State \(\to\) Set[State]
  (define (step \(\mstate\))
    (match \(\mstate\)
      [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
       (match \(\mexp\)
         [(var\(\superscript\mlab\) \(\mvar\))
          (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
         [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
         [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
         [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
      [(co \(\msto\) \(\mkont\) \(\mval\))
       (match \(\mkont\)
         ['mt (set (ans \(\msto\) \(\mval\)))]
         [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
         [(fn\(\superscript\mlab\) \(\mval'\))
          (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
            (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
         [(fi\(\superscript\mlab\) c a \(\rho\))
          (for/set ((k (get-cont \(\msto\) l)))
            (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
      [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
       (match fun
         [(clos l \(\mvar\) \(\mexp\) \(\rho\))
          (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
          (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
         [(op \(\mop\))
          (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
                     (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [_ (set)]))]))
\end{alltt}
\caption{Implementation of machine transition relation.}
\end{figure}

@@

\begin{alltt}
;; Expr \(\to\) Set[State]
(define (eval e)
  (fix step (set (inj e))))
\end{alltt}

@@

\begin{alltt}
;; \(\forall\) (X) ((X \(\to\) Set[X]) \(\to\) (Set[X] \(\to\) Set[X]))
(define ((appl f) s)
  (for/fold ([i (set)])
    ([x (in-set s)])
    (set-union i (f x))))

;; Calculate fixpoint of (appl f).
;; \(\forall\) (X) ((X \(\to\) Set[X]) Set[X] \(\to\) Set[X])
(define (fix f s)
  (let loop ((accum (set)) (front s))
    (if (set-empty? front)
        accum
        (let ((new-front ((appl f) front)))
          (loop (set-union accum front)
                (set-subtract new-front accum))))))
\end{alltt}

@@

\begin{alltt}
;; \(\widehat{State}\) = (cons (Set Conf) Store)

;; (State \(\to\) Set[State]) \(\to\) \(\widehat{State}\) \(\to\) \(\widehat{State}\)
(define ((wide-step step) state)
  (match state
    [(cons cs \(\msto\))
     (define ss ((appl step)
                 (for/set ([c cs]) (c->s c \(\msto\)))))
     (set (cons (for/set ([s ss]) (s->c s))
                (join-stores ss)))]))
\end{alltt}

@@

\begin{align*}
force(\clos{\mvar,\mexp,\menv},\msto) &= \{\clos{\mvar,\mexp,\menv}\}\\
force(\saddr\maddralt,\msto) &= \msto(\maddralt)\\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\; 
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},v,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}} \\
\mbox{where } &\mval' \in force(\mval,\msto) \\
bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], \msto\sqcup[\maddr \mapsto force(\mval, \msto)], \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

The force for the second rule is necessary to keep the same type for
the \(\kfn{}\) continuation, but we would rather use an address. This
introduces (acceptably) more imprecision, but has the same precision as an ANF
analysis\footnote{Exercise for the reader.}.

Lazy:
   cpu time: 32481 real time: 32881 gc time: 547

@@

\begin{figure}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \compile{\mexp}(\epsilon,\varnothing,\varnothing,\kmt) \multimachstep \mstate \} \text{ where }
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mcomp,\menv,\maddr},\mval,\msto} & \machstep
\mcomp^\mcntr(\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr})
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\strue,\msto} & \machstep
\mcompi0^\mcntr(\menv,\msto,\mkont)
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mcompi0,\mcompi1,\menv,\maddr},\sfalse,\msto} & \machstep
\mcompi1^\mcntr(\menv,\msto,\mkont)
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mcomp,\menv},\mval,\msto,\mkont} & \machstep
\mcomp^{\mcntr'}(\menv',\msto',\mkont) \\
\text{ where }\menv',\msto',\mcntr' &= \mathit{bind\,}^{ \mcntr}_\mlab(\msto,\mvar,\mval)
\\
\ap{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto} \\
\text{ where }\mkont &\in\msto(\maddr)
\text{ and } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for compiled ISWIM}
\label{fig:caam}
\end{figure}

@@

\begin{alltt}
;; \(\widehat{State} \to \widehat{State}\)
(define (wide-step-specialized state)
  (match state
    [(cons \(\msto\) cs)
     (define-values (cs* \(\msto\)*)
       (for/fold ([cs* \(\varnothing\)] [\(\msto\)* \(\msto\)])
         ([c cs])
         (match (step-compiled^ (cons \(\msto\) c))
           [(cons \(\msto\)** cs**)
            (values (\(\cup\) cs* cs**) (\(\sqcup\)-store \(\msto\)* \(\msto\)**))])))
     (cons \(\msto\)* (\(\cup\) cs cs*))]))
\end{alltt}
%% Should be commutted to beginning

@@

\begin{alltt}
;; \(\widehat{State} \to \widehat{State}\)
(define (wide-step-specialized state)
  (match state
    [(cons \(\msto\) cs)
     (define-values (cs* \(\msdiff\))
       (for/fold ([cs* (set)] [\(\msdiff\)* '()])
         ([c cs])
         (match (step-compiled^ (cons \(\msto\) c))
           [(cons \(\msdiff\)** cs**)
            (values (\(\cup\) cs* cs**) (++ \(\msdiff\)** \(\msdiff\)*))])))
     (cons (update \(\msdiff\) \(\msto\)) (\(\cup\) cs cs*))]))
\end{alltt}

@@

\begin{alltt}
 (let ((id (\(\lambda\) (x) x)))
   (begin (id 1) (id 2)))
\end{alltt}
